---
title: "STAT/CSSS 564: Assignment 2 Solutions"
author: Dongsheng Dong
output: html_document
---

Put R packages used in a single chunk:
```{r}
rm(list = ls())
library("rstan")
set.seed(042317)

# Load data
data = read.csv("pr2data.csv", header = T)
```


## How much does the prior influence the posterior?


### Q1
A class of prior densities is conjugate for a sampling model $p(y_1, . . . , y_n|θ)$ if the posterior distribution is also in the class. In other words, if the prior and posterior distribution belong to the same distribution family, then the prior distribution is called *conjugate prior*. The advantage of using a conjugate prior is that it make the estimation of posterior distribution a little bit simpler. In other words, if we know the distribution of prior, then we know what the approxiate distribution of posterior and the formula for normalizing the posterier. In that case, the posterior distribution could be approximated with the Monte Carlo method or any other methods. The main disadvantage of conjugate prior is that it is not flexible enough to capture all possible forms of prior information. 

Suppose the prior distribution is non-conjugate such as the Laplace distribution, then we wouldn't be able to assume the distribution of the posterier directly. But we know:
$$ f(\mu) = \frac {1} {\sqrt{2 \sigma_0^2 \pi}} e^{\frac{-(x-\mu_0)^2}{2\sigma_0^2}}$$
$$ f(x|\mu) = \frac {1} {\sqrt{2 \sigma^2 \pi}} e^{\frac{-(x-\mu)^2}{2\sigma^2}}$$
$$f(x|\mu, b) = \frac{1}{2b} exp(- \frac{|x-\mu|}{b})$$

So we know that posterier distribtion with a Laplace prior is
$$f(\mu|b, x) \propto \frac{1}{2b} exp(- \frac{|x-\mu|}{b}) * \frac {1} {\sqrt{2 \sigma_0^2 \pi}} e^{\frac{-(x-\mu_0)^2}{2\sigma_0^2}}$$
Finally, a normal prior may be preferred because we then know the distribution of posterier is normal distribution. In this case, we could use the probability density function (PDF) of normal distribution to calculate and normalize the posterier distribution, which makes the estimation and interpretation of posterier much easier. 

### Q2
####Lots of data vs. not much data
From the question we know that the variance is inversely propritional to the amount of data. The more data we have, the less variance there. When the data goes to infinite, the distribution of posterier will becomes narrower and tighter as the variance become much smaller. However, if we only have a few data points, the estimation of the posterier become less precise so we may not be able to draw a valid inference about $\mu$. 


####Strong prior information vs. weak prior information
Strong prior information would dominate the information contained in the data being analyzed and thus has an strong influence on the posterier distribution. In this case, the posterier distribution may be led towards the prior. Weak prior information is less dominant and thus it allows the data speak, therefore, the results is more data-driven.

####Lots vs. little noise in the data
If there are a lot of noises in the data, it increases the variance and thus cause the posterer distribution to be wider. However, if there are little noise noise in the data, the posterier will become tighter given its small variance.


####Data that is incompatible with the prior vs. data that is compatible with the prior
If the data is compatible with the prior, then the variance should be reduced and the estimation of mean should be more likely to be close to the true mean in terms of the posterier distribution. Therefore, we should have a much narrower posterier distribution which should provide valid inference about $\mu$. Contrarily, if the data is not compatible with prior, then there are much more noise in the data which increase variance and thus will lead to a much wider distribution of posterier and we may not have valid inference about $\mu$. 

### Q3
```{r 3}
# linear model with intercept only
m1 = lm(x~1, data = data)
# estimated mean
m1
# estimated variance
summary(m1)$sigma^2
# 97% CI
confint(m1, level = 0.97)
# Result results
cat("Mean:", 10) 
cat("Variance:", 104.31) 
cat("97% CI: [9.30, 10.70]")
```

A linear model was fitted with just an intercept. Since it is an intercept-only model, the estiamted mean is equal to the coefficient of the intercept, which is 10. The variance is 104.3064. The centered 97% confidence intereval for $\mu$ is [9.30, 10.70], which means: under the assumed model, 95% of the time the estiamted confidence intervals capture the true $\mu$ or the probability that $\mu$ lies between 9.30 and 10.70 is 0.95.

### Q4
```{r 4}
# list all data needed for m2
m2_data = list(
  y = data$x,
  N = nrow(data)
)
# fit the bayesian model
m2 = stan("lm_q4.stan", data = m2_data, iter = 1000, chains = 2)
m2
# 97% CI
print(m2, prob = c(0.005, 0.985))
# Result results
cat("Mean:", 10.01) 
cat("Variance:", 0.00) # very close to 0.01
cat("97% CI: [9.76, 10.19]")
```

I would characterize this prior as highly informative because it provides very specific and relatively precise information about the variable so that we know $\mu$ follows a normal distribution with mean of 10 and variance of 0.01. Note the variance is very small rather than some general numbers like 10000. Give the small variance provided, I belieive this prior is highly informative. The estiamted mean is 10.1. The variance is 0.00. The centered 97% confidence intereval for $\mu$ is [9.76, 10.19], which means: under the assumed model, the probability that $\mu$ lies between 9.76 and 10.19 is 0.95.

### Q5
```{r 5}
# generate 10000 independent observations from a N(10, 100) distribution
data_q5 = rnorm(10000, mean = 10, sd = 10)
data_q5 = as.data.frame(data_q5)
names(data_q5) = "x"
# concatenate new data to the old data
data_cat = rbind(data, data_q5)

# fit standard linear model with intercet only
m3 = lm(x ~ 1, data = data_cat)
# estimated mean
m3
# estiamte variance
summary(m3)$sigma^2
# 97% CI
confint(m3, level = 0.97)
# Result results
cat("Standard lm mean:", 10.19) 
cat("Standard lm variance:", 100.36) 
cat("Standard lm 97% CI: [9.98, 10.40]")


# fit a bayesian model with normal likelihood and an improper flat prior
# list all data needed for m2
m4_data = list(
  y = data_cat$x,
  N = nrow(data_cat)
)
# fit the bayesian model
m4 = stan("lm_q5.stan", data = m4_data, iter = 1000, chains = 2)
m4
# 97% CI
print(m4, prob = c(0.005, 0.985))
# Result results
cat("Bayesian mean:", 10.19) 
cat("Bayesian variance:", 0.00) # very close to 0.01
cat("Bayesian 97% CI: [9.91, 10.38]")
```

### Q6
In Q4 we fitted a bayesian model with a highly informative prior ($N(10, 0.1^2)$) with a few observations (N = 10000), while in Q5 we fitted another bayesian model with a improper flat prior with a much larger sample size (N = 11000). Results show that the means, variances and the confidence intervals for the two obtained posterier distributions are very close. That indicates that when the prior is uninformative, adding more sample sizes could still lead to a valid estiamtion of posterier disribution. Linking back to the formula for the posterier distribution $\mu|x$ given before Q1, the variance is inverse proportional to the amount of data. The more data we have, the less variance in the model. Therefore, the inferrence could still be valid even though we have an uninformative prior as long as we have sufficient data.

### Q7
Two simplifying assumptions were made in this problem. The first assumption is that we assume the variance is known and it is equal to a constant number (variance is equal to 100). The second assumption is that the additional data are centered to have a mean of 10, which is rarely known and observed to us in real-life sitatuions. If we relax those assumptions, then the variance has to be estimated as well and the standard error of mean is going to change when estimating mean, which may change the distribution of the posterier.
